A Study in Pairwise Testing Heuristics for Granger-causal Network Estimation

### 2019-09-28 paper/paper.tex
Working on presentation.

\title{Graph Topological Aspects of Granger Causal Network Learning }
\author{R. J. Kinnear \\
  \small\href{mailto:ryan@kinnear.ca}{ryan@kinnear.ca} \\
  \small\url{https://github.com/RJTK} \and R. R. Mazumdar \\
  \small\href{mailto:mazum@uwaterloo.ca}{mazum@uwaterloo.ca} }

@@

where $c(t)$ is a purely deterministic sequence\footnote{the purely
  deterministic sequence $c(t)$ is one which lies in the remote past
  $\bigcap_{\tau=1}^\infty \H_{t - \tau}^x$ of the process.  For such
  processes a single sample $c(t_0)$ is enough to determine $c(t)$ for
  every $t$.  For example,
  $c(t) = \text{sin}(2\pi t + \Theta);\; \Theta \sim \mathcal{U}[-\pi,
  \pi]$}, $v(t)$ is an uncorrelated sequence and $A(0) = I$.  We will
assume that $c(t) = 0$, which in practice is to say that the process
has been detrended.  We additionally require that this representation
can be inverted to yield the $\VAR(\infty)$ form

>>>>

\title{Graph Topological Aspects of Granger Causal Network Learning}
\author{R. J. Kinnear \\
  \small\href{mailto:ryan@kinnear.ca}{ryan@kinnear.ca} \\
  \small\url{https://github.com/RJTK} \and R. R. Mazumdar \\
  \small\href{mailto:mazum@uwaterloo.ca}{mazum@uwaterloo.ca} }

@@

where $c(t)$ is a purely deterministic sequence\footnote{the purely
  deterministic sequence $c(t)$ is one which lies in the remote past
  $\bigcap_{\tau=1}^\infty \H_{t - \tau}^x$ of the process.  For such
  processes a single sample $c(t_0)$ is enough to determine $c(t)$ for
  every $t$.  For example,
  $c(t) = \text{sin}(2\pi t + \Theta);\; \Theta \sim \mathcal{U}[-\pi,
  \pi]$}, $v(t)$ is an uncorrelated sequence and $A(0) = I$.  We will
assume that $c(t) = 0$, which in practice is to say that the process
has been detrended.  We additionally require that this representation
tcan be inverted to yield the $\VAR(\infty)$ form

### 2019-09-28 paper_appl/paper_appl.tex
Working on presentation.

>>>>

\section{Laten Structure Recovery Performance Metric}
The Granger causal graph structure described in section
\ref{sec:theory} is entirely latent.  That is, we can never observe
the true causality graph (assuming such a structure \textit{exists}),
which makes evaluating different support recovery algorithms
difficult.  Broadly speaking, most work on the topic evaluates the
performance of competing causality graph recovery algorithms in one of
two ways (or both): (1) theoretical proofs or (2) simulations where
the true graph is known by construction.  In this work our goal is to
establish as valid a third possibility: the performance of a second
step classifier.

\subsection{Introduction and Motivation}
More precisely, we suppose we have a two distinct data generating
processes (``subjects'') labeled $S^{(1)}$ and $S^{(2)}$.  Each
subject has their own fixed, distinct, and unobservable Granger
causality graph $\gcg^{(i)}$.  And, when queried, the subject
returns% \footnote{Formally, we could define a distribution $t_0$ on
  % $\N$ and take $X^{(i)} = \{x^{(i)}(t)\}_{t = t_0}^{t_0 + T}$.  Since
  % $x(t)$ is stationary, this doesn't effect the statistics of the
  % sample}
a finite length $T$ sample of time series data
$X^{(i)} \defeq \{x^{(i)}(t)\}_{t = 1}^{T}$ from a stationary process
whose temporal dependence structure corresponds to the causality graph
structure of $\gcg^{(i)}$.  We suppose that we collect $N$ independent
samples $\{X^{(i)}_n\}_{n = 1}^N$ from each subject.

In addition to the subjects, we have $K$ algorithms $A_1, \ldots, A_K$
which consume such a dataset $X_n^{(i)}$ and return an estimate
$\widehat{\gcg}_{k,n}^{(i)}$ of graph $\gcg^{(i)}$.  That is, if we
view graph estimates as points in $\{0, 1\}^{n^2}$ we have
$A_k: \R^{T \times n} \rightarrow \{0, 1\}^{n^2}$.

Applying each algorithm to each sample of time series data allows us
to construct $K$ datasets $\mathcal{D}_1, \ldots, \mathcal{D}_K$, each
of which contains $2N$ labeled graph estimates
$(\widehat{\gcg}_{k, n}^{(i)}, i)$:

\[
  \mathcal{D}_k = \bigcup_{i \in \{1, 2\}} \bigcup_{n = 1}^N \{(\widehat{\gcg}_{k, n}^{(i)}, i)\}.
\]

Finally, we fit a classifier
$C_k: \{0, 1\}^{n^2} \rightarrow \{1, 2\}$ to the dataset
$\mathcal{D}_k$ which consumes an estimated causality graph and
attempts to predict the subject label $i \in \{1, 2\}$ that generated
the data from which the graph was estimated.  If classifier $C_{k_1}$
has superior out-of-sample test accuracy in comparison to $C_{k_2}$,
we consider this as evidence that Algorithm $A_{k_1}$ has made more
accurate estimates of $\gcg^{(i)}$ than has Algorithm $A_{k_2}$.  That
is, if for new graph estimates
$\widehat{\gcg}_{k_1}^{(i)}, \widehat{\gcg}_{k_2}^{(i)}$ we have

\begin{equation}
  \label{eqn:comparison_metric}
  \P\big\{C_{k_1}(\widehat{\gcg}_{k_1}^{(i)}) = i\big\} > \P\big\{C_{k_2}(\widehat{\gcg}_{k_2}^{(i)}) = i\big\}.
\end{equation}

\begin{remark}
We will provide mathematical justification for this evaluation
strategy below, but some remarks are in order.

Firstly, if Equation \ref{eqn:comparison_metric} holds, it does not
constitute \textit{proof} that $A_{k_1}$ is superior to $A_{k_2}$ for
the task of estimating Granger causality graphs.  Indeed, it would be
possible to construct an algorithm $A$ which attempts to encode the
idiosyncracies of $s^{(i)}$ present in the data $x^{(i)}(t)$, making
no attempt at actually recovering $\gcg^{(i)}$.  We therefore
formulate our theoretical justification such that we make no reference
to the actual recovery algorithm.

Secondly, we claim that this evaluation strategy is distinct from the
above evaluation metrics (1) and (2) (i.e. mathematical proof, and
simulation), but we do not claim it is \textit{better}.  Merely, we
propose it as a third mechanism for comparing different recovery
algorithms that may provide further insight and quantify performance
along another axis.
\end{remark}

\subsection{Theoretical Analysis}
